Parallel global optimization on GPU \thanks{The research is supported by the grant of the Ministry of education and science of the Russian Federation (the agreement of August 27, 2013 No 02.B.49.21.0003)}

### 2019-10-07 2016/JoGO_GPU/BarkalovGergel.tex
JOGO2016 paper second version

\begin{abstract}
This work considers a parallel algorithm for solving multidimensional multiextremal optimization problems. This algorithm uses Peano-type space filling curves for dimension reduction. Conditions of non-redundant parallelization of the algorithm are considered. Implementation of the algorithm on modern computing systems with the use of a graphics processing unit (GPU) is investigated. Speedup of the algorithm using GPU as compared with the same algorithm implemented on CPU only is demonstrated experimentally. Computing experiments are carried out on a series of several hundred multidimensional multiextremal problems.
\keywords{Global optimization \and Parallel computing \and Graphics processing unit \and Peano-type space filling curves \and Characteristical algorithms}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

This paper considers parallel algorithms for solving multiextremal optimization problems. The objective function in such problems is characterized by existence of several local optima (usually their number is not known and can be large). In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of an a priori information about the function, allowing us to link possible values of an optimizable function with known values at search points. Very often, such information about the problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with an a priori unknown constant $L$. The objective function can be represented by the algorithm for calculating its values in the search domain, i.e. it can be a ``black-box'' function. Similar problems are common in applications (problems of optimal design of objects and technological processes in various fields of technology, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

Multiextremal optimization problems offer little in terms of analytical solution, while numerical solution is computationally challenging as computational costs often grow exponentially in dimension. By employing modern parallel computational systems it is possible to broaden the area of applicability of global optimization methods. At the same time, the problem of efficient parallelization of the global search process arises. Therefore, development of efficient parallel methods for numerical solution of multiextremal optimization problems and their software implementation for modern computational systems is the obvious focus of research in this area.

One of the promising trends in the field of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

However, in the field of global optimization the potential of GPUs is not completely utilized. GPUs are generally used for parallelization of algorithms, which one way or another are based on the concept of random search (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi}); a review of this trend of GPU optimization algorithms is given in \cite{RefLangdon}. Owing to their stochastic nature, algorithms of this type generally do not guarantee convergence to a globally optimal solution, and that sets them apart from deterministic methods.

Some parallel variants are proposed for many deterministic algorithms of lipschitzian global optimization with guaranteed convergence \cite{RefGergel2005,RefEvtushenko,RefHe,RefPaulavicius}. However, these versions of algorithms are parallelized on CPU using MPI and/or OpenMP technologies; there are no implementations on GPU. The present article contains the results of an investigation into the GPU implementation of the global search parallel algorithm developed according to the information-statistical approach presented in the monograph \cite{RefStrongin2000}.

The main part of the paper has the following structure. Section \ref{sec:2} contains a description of the basic algorithm of global search. Section \ref{sec:3} contains a scheme of its parallelization, and section \ref{sec:4} includes theoretical statements that characterize the speedup and non-redundancy of the parallel algorithm. In section \ref{sec:5}, ways of possible implementation of the parallel algorithm on GPU are discussed. Section \ref{sec:6} contains the results of numerical experiments; a comparison of the sequential algorithm and parallel CPU and GPU versions for solving a series of test multidimensional multiextremal problems is also made. Section \ref{sec:7} concludes the paper.

@@

Let us consider the problem of search for a global minimum of an $N$-dimensional function $\varphi(y)$ within a hyper interval $D$
\begin{eqnarray}\label{eq:1}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}

Let us assume that the function $\varphi$ satisfies a Lipschitz condition with an a priori unknown constant $L$
\[
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D,\; 0<L<\infty.
\]

There is a number of ways to adapt effective one-dimensional algorithms for solving multidimensional problems; see, for example, the diagonal partitions method in \cite{RefSergeyev2006} or the simplicial partitions method in \cite{RefZilinskas}. In this paper, we will use the approach based on the idea of dimension reduction by means of a Peano curve $y(x)$, which continuously and unambiguously maps the unit interval [0,1] onto the $n$-dimensional cube
\[
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{y(x):0\leq x \leq 1 \right\}.
\]
Problems of numerical construction of Peano-type space filling curves and the corresponding theory are considered in detail in \cite{RefStrongin2000}, \cite{RefSergeyev2013}. Here we will note that a numerically constructed curve (\textit{evolvent}) is an approximation to a theoretical Peano curve with an accuracy $2^{-m}$ , where $m$ is an evolvent construction parameter. It means that using an evolvent we can distinguish points in a multidimensional space, the coordinates of which differ by not less than $2^{-m}$; closer points are considered as coincident. Thereby an evolvent constructed with a fixed $m$ allows obtaining a solution of a problem with an accuracy of not more than $2^{-m}$ by coordinate.

By using this kind of mapping it is possible to reduce the multidimensional problem~(\ref{eq:1}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}}.
\]
An important property of such mapping is preservation of boundedness of function relative differences  (see \cite{RefStrongin2000}): if the function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, then the function $\varphi(y(x))$ on the interval $[0,1]$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where the H{\"o}lder constant $H$ is linked to the Lipschitz constant $L$ by the relation
\[
H=4Ld\sqrt{N},\;\; d=\max{\left\{b_i-a_i:1\leq i \leq N\right\}}.
\]

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying the H{\"o}lder condition.

The considered algorithm for solving this problem (here, according to \cite{RefStrongin2000}) involves constructing a sequence of points $x^k$, where the values of minimized function $z^k = f(x^k)$ are calculated. Let us call the process of calculating the function value (including the construction of an image $y^k=y(x^k)$) the ``trial'', and the pair $(x^k, z^k)$, the ``trial result''. The set of pairs $\left\{(x^k, z^k)\right\}, 1\leq k\leq n,$ makes up the search data collected using the method after carrying out $n$ steps. The rules that determine the work of the global search algorithm are as follows.

At the first iteration of the method the trial is carried out at an arbitrary internal point $x^1$ of the interval $[0,1]$. The point of trial at the next iteration $(k+1)$ is determined according to these rules.

Step 1. Renumber points of the set
\[
X_k=\{x^1,\dots,x^k\}\cup\left\{0\right\}\cup\left\{1\right\},
\]
which includes boundary points of the interval $[0,1]$ and the points of the previous trials, with subscripts in increasing order of coordinate values, i.e.
\[
0=x_0<x_1<\dots <x_k<x_{k+1}=1.
\]

@@

Step 3. Calculate a \textit{characteristic} for every interval $(x_{i-1}, x_i), \; 1\leq i \leq k+1,$   according to the following formulas
\[
R(1)=2\Delta_1-4\frac{z_1}{M},
\]
\begin{equation}\label{eq:14}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\frac{z_i+z_{i-1}}{M},1<i<k+1,
\end{equation}
\[
R(k+1)=2\Delta_{k+1}-4\frac{z_k}{M}.
\]

Step 4. Find interval $(x_{t-1},x_t)$ with the maximum characteristic
\[
R(t)=\max{\left\{R(i): 1 \leq i \leq k+1\right\}}.
\]

Step 5. Carry out a trial at the point $x^{k+1}\in(x_{t-1},x_t)$, calculated using the following formulas
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2}, \textrm{ if } t=1 \textrm{ or } t=k+1,
\]
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t<k+1.
\]

The algorithm terminates if the condition $\Delta_t<\epsilon$ is satisfied; here $\epsilon>0$ is the preset accuracy. For estimation of the global solution, values
\[
f_k^\ast=\min_{1\leq i \leq k}f(x^i), \ x_k^\ast=\arg \min_{1\leq i \leq k}f(x^i).
\]
are selected.

Remark 1. The normalized characteristics $R(i)$ from (\ref{eq:14}) can be viewed as probabilities of locating a global minimum within the interval $(x_{i-1},x_i), 1\leq i \leq k+1$. This statement is proved in \cite{RefStrongin2000}. Thus, at each iteration a new trial point is selected inside the interval, which has the greatest probability of finding a global minimum.

Remark 2. The proof of convergence of this algorithm is provided in \cite{RefStrongin2000}. The modifications taking into account existence of inequality constraints in the problem and the information about the objective function derivative are given in \cite{RefBarkalov,RefGergel1996,RefGergel1997}.

Let us illustrate the work of this \textit{global search algorithm} (GSA) using a dimension reduction for minimization of a two-dimensional multiextremal function
\begin{eqnarray} \nonumber \label{eq:19}
\varphi(y)= -&\left\{\left(\sum^{7}_{i=1}\sum^{7}_{j=1}A_{ij}g_{ij}(y)+B_{ij}h_{ij}(y)\right)^2+\right. \\
&\left.\left(\sum^{7}_{i=1}\sum^{7}_{j=1}C_{ij}g_{ij}(y)+D_{ij}h_{ij}(y)\right)^2\right\}^{1/2},\\ \nonumber
\end{eqnarray}
where
\begin{eqnarray} \nonumber
& y=(y_1,y_2)\in R^2, 0 \leq y_s \leq 1, s=1,2, \\ \nonumber
& g_{ij}(y)=\sin(i\pi y_1)\sin(j\pi y_2),  \\ \nonumber
& h_{ij}(y)=\cos(i\pi y_1)\cos(j\pi y_2), \nonumber 
\end{eqnarray}
and coefficients $A_{ij}, B_{ij}, C_{ij}, D_{ij}$  are taken uniformly in the interval $[-1,1]$. The experiment used the method parameters $r=3$, $\epsilon=10^{-3}$, and the evolvent construction parameter $m=12$. Fig.~\ref{fig:1} shows the contour lines of a function generated according to (\ref{eq:19}) and the points of 1103 trials carried out using the method before the required accuracy was obtained.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig1.jpg} 
  \caption{Minimization of a two-dimensional function by the sequential algorithm}
  \label{fig:1}       % Give a unique label
\end{figure}

@@

The first option is to parallelize calculation of the target function describing the optimized object. This method can provide some speedup, but is specific for each problem.

The second option is to parallelize implementation of algorithm computing rules for selection of the next trial point. In this case the way of parallelization will depend on the class of algorithms. Moreover, these rules are often rather simple and it is inexpedient to parallelize them (overhead costs of the organization of parallelism will reduce possible speedup to zero).

The third option is to  change the scheme of an algorithm for the purpose of carrying out several trials in parallel. This approach is most promising since it is characterized by efficiency (the part of computing process, in which the bulk of calculations is carried out, is parallelized) and generality (it is applicable for a wide class of characteristical algorithms of multiextremal optimization \cite{RefGrishagin1997}).

Let us describe a change of rules of global search algorithm for the purpose of simultaneous (synchronous) carrying out in one iteration of method $p$ of trials. Let us make $k(n)$ the total number of trials, carried out after $n$ parallel iterations.

Suppose $n\geq 1$  iterations of the method are carried out, in which trials were performed at $k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next $n+1$ iteration are determined according to the rules.

Steps 1--3 of the parallel algorithm fully repeat steps 1--3 of the sequential global search algorithm.

@@

Step 5. Carry out new trials at points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using the formulas
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2}, \textrm{ if } t_j=1 \textrm{ or } t_j=k+1,
\]
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t_j<k+1.
\]

The algorithm terminates if the condition $\Delta_{t_j}<\epsilon$ is satisfied at least for one number $t_j, 1 \leq j \leq p$ ; $\epsilon>0$ is the preset accuracy.  For estimation of the global solution, values
\[
f_k^\ast=\min_{1\leq i \leq k}f(x^i), \ x_k^\ast=\arg \min_{1\leq i \leq k}f(x^i).
\]
are selected.

The modifications of the parallel algorithm with the account of the information about the minimized function derivative are given in \cite{RefGergel1999}.

This method of organizing parallel computing has the following justification \cite{RefStrongin2000,RefGrishagin1997}. The characteristics of intervals (\ref{eq:14}) used in the global search algorithm can be considered as probability measures of the global minimum point location in these intervals. Inequalities (\ref{eq:21}) arrange intervals according to their characteristics, and trials are carried out in parallel in the first $p$ intervals with the largest probabilities.

Let us use a parallel algorithm of global search for solving a problem from section \ref{sec:2} with the numbers of parallel trials $p=2$ and $p=4$. With $p=2$ the number of iterations was $569$, and the number of trials was $1138$. With $p=4$ the number of iterations was reduced to $270$, and $1080$ trials were carried out. Fig.~\ref{fig:2} shows contour lines of the function with points of the trials performed by the method with $p=2$. Points of redundant trials carried out using a parallel algorithm are marked in the figure by "+". The concept of redundancy is discussed in more detail in the following section.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig2.jpg} 
  \caption{Minimization of a two-dimensional function by the parallel algorithm}
  \label{fig:2}
\end{figure}

@@

Let us describe theoretical properties of a parallel algorithm, which characterize its speedup. One of the main indicators of efficiency of parallel algorithms (in any field, not only in global optimization) is a speedup in time
\[
S(p)=T(1)/T(p),
\]
where $T(1)$ is the time required to solve a problem by a sequential algorithm, and $T(p)$ is the time for solving the same problem by a parallel algorithm in a system with $p$ computing elements. The characteristic of efficiency of a parallel algorithms (in relation to optimization algorithms) is also a speedup in iterations
\begin{equation}\label{eq:26}
s(p)=n(1)p/n(p),
\end{equation}
where $n(1)$ is the number of trials carried out using the sequential method, and $n(p)$ is the number of trials carried out using the parallel method with $p$ processors. This characteristic is especially important since for solving applied problems the time required to carry out a trial exceeds the time required to process its results.

It is obvious that the number of trials $n(p)$ for sequential and parallel algorithms described in sections \ref{sec:2} and \ref{sec:3} will differ. Actually, the sequential algorithm of global search when selecting point $x^{k+1}$ of the next $k+1$ trial possesses complete information received at the previous $k$ iterations. Based on the same information, the parallel algorithm of global search selects $p$ points $x^{k+j}, 1\leq j \leq p$, rather than one at iteration $k+1$. It means that selection of point $x^{k+j}$ is carried out in absence of information on the trial results at points $x^{k+i}, 1\leq i<j$. Only the first point $x^{k+1}$ will coincide with the point selected by the sequential algorithm. Other trials points generally cannot coincide with the points generated by the sequential algorithm. The efficiency offered by the use of parallel processors can be reduced when such trials are performed. Therefore, let us consider such trials as ``redundant'', and the value 
\[
\lambda(p) = \left\{
   \begin{array}{lr}
     (n(p)-n(1))/n(p), & n(p) > n(1),\\
     0, & n(p)\leq n(1),
   \end{array}
\right.
\]
will represent ``method redundancy''.

Let us set a series of trials $\left\{x^k\right\}$ and $\left\{y^m\right\}$ generated correspondingly by sequential and parallel algorithms for solving the same problem with $\epsilon = 0$ in the stopping condition. The following theorem from \cite{RefStrongin2000} determines the number of computing elements $p$, which can be involved for non-redundant parallelization.

>>>>

\begin{abstract}
This work considers a parallel algorithm of solution of multidimensional multiextremal optimization problems. This algorithm uses Peano-type space filling curves for dimension reduction. Conditions of non-redundant parallelization of the algorithm is considered. The issue of implementation of the algorithm on modern computing systems with use of a graphics processing unit (GPU) is investigated. Speedup of the algorithm using GPU as compared with the same algorithm implemented on CPU only is demonstrated experimentally. Computing experiments are carried out on a series of several hundred of multidimensional multiextremal problems.
\keywords{Global optimization \and Parallel computing \and Graphics processing unit \and Peano-type space filling curves \and Characteristical algorithms}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

This present paper considers parallel algorithms for the solution of problems of multiextremal optimization. In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of a priori information about the function, allowing us to link possible values of an optimizable function with known values in search points. Very often, such information about the solvable problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with a priori unknown constant $L$. The objective function can be represented by the algorithm of calculation of its values in the search domain, i.e. it can be a ``black-box'' function. Similar problems are widespread in applications (problems of optimal design of objects and technological processes in various technical domains, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

Multiextremal optimization problems offer little in terms of analytical solution and numerical solution is computationally challenging as computational costs often grow exponentially in dimension. Employing modern parallel computational systems broadens the area of applicability of global optimization methods and, at the same time, poses a problem of efficient parallelization of the global search process. Therefore development of efficient parallel methods for numerical solution of multiextremal optimization problems and their software implementation for modern computational systems is the main direction of research in this area.

One of the promising directions of development in the domain of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years, and has become more programmable. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

However, in the domain of global optimization the potential of GPUs is not completely unlocked. GPUs are generally used for parallelization of algorithms, which one way or another are based on the concept of random search (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi}); a review of this direction of GPU optimization algorithms is given in \cite{RefLangdon}. Owing to their stochastic nature, algorithms of this type generally do not guarantee convergence to a globally optimal solution, and that sets them apart from deterministic methods.

For many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence parallel variants are proposed \cite{RefGergel2005,RefEvtushenko,RefHe,RefPaulavicius}. However, these versions of algorithms are parallelized on CPU using MPI and/or OpenMP technologies; there are no implementations on GPU. The present article contains the results of an investigation into the GPU implementation of the global search parallel algorithm developed according to the information-statistical approach presented in the monograph \cite{RefStrongin2000}.

The main part of the paper has the following structure. Section \ref{sec:2} contains a description of the basic algorithm of global search. Section \ref{sec:3} contains a scheme of its parallelization, and section \ref{sec:4} includes theoretical statements that characterize speedup and non-redundancy of the parallel algorithm. In section \ref{sec:5}, ways of possible implementation of parallel algorithm on GPU are discussed. Section \ref{sec:6} contains the results of numerical experiments; comparison of the sequential algorithm and parallel CPU and GPU versions for the solution of a series of test multidimensional multiextremal problems is also made. Section \ref{sec:7} concludes the paper.

@@

Let us consider the problem of search for a global minimum of an $N$-dimensional function $\varphi(y)$ within hyper interval $D$
\begin{eqnarray}\label{eq:1}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}

Let us assume that function $\varphi$ satisfies a Lipschitz condition with a priori unknown constant $L$
\[
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D,\; 0<L<\infty.
\]

There is a number of ways to adapt effective one-dimensional algorithms for solution of multidimensional problems; see, for example, diagonal partitions method in \cite{RefSergeyev2006} or simplicial partitions method in \cite{RefZilinskas}. In this paper, we will use the approach based on the idea of dimension reduction by means of a Peano curve $y(x)$, which continuously and unambiguously maps the unit interval [0,1] onto the $n$-dimensional cube
\[
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{y(x):0\leq x \leq 1 \right\}.
\]
Problems of numerical construction of Peano-type space filling curves and the corresponding theory are considered in detail in \cite{RefStrongin2000}, \cite{RefSergeyev2013}. Here we will note that a numerically constructed curve is an approximation to a theoretical Peano curve with accuracy $2^{-m}$ , where $m$ is a curve construction parameter.

Use of mapping of this kind allows a reduction of a multidimensional problem~(\ref{eq:1}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}}.
\]
An important property is preservation of boundedness of function relative differences  (see \cite{RefStrongin2000}): if function $\varphi(y)$ in domain $D$ satisfies Lipschitz condition, then function $\varphi(y(x))$ on interval $[0,1]$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where H{\"o}lder constant $H$ is linked to Lipschitz constant $L$ by the relation
\[
H=4Ld\sqrt{N},\;\; d=\max{\left\{b_i-a_i:1\leq i \leq N\right\}}.
\]

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying H{\"o}lder condition.

The considered algorithm of the solution of this problem (here, according to \cite{RefStrongin2000}) presupposes construction of sequence of points $x^k$, where values of minimized function $z^k = f(x^k)$ are calculated. Let us call the process of calculation of the value of function (including construction of an image $y^k=y(x^k)$) ``trial'', and pair $(x^k, z^k)$ --  the ``trial result''. Set of pairs $\left\{(x^k, z^k)\right\}, 1\leq k\leq n,$ make up the search data collected using the method after carrying out $n$ steps. The rules that determine the work of the global search algorithm are as follows.

At the first iteration of the method the trial is carried out in arbitrary internal point $x^1$ of interval $[0,1]$. The point of trial at the next iteration $(k+1)$ is determined according to these rules.

Step 1. Renumber points of the set
\[
X_k=\{x^1,\dots,x^k\}\cup\left\{0\right\}\cup\left\{1\right\},
\]
which includes boundary points of interval $[0,1]$ and points of the previous trials, with subscripts in increasing order of coordinate values, i.e.
\[
0=x_0<x_1<\dots <x_k<x_{k+1}=1.
\]

@@

Step 3. Calculate \textit{characteristic} for every interval $(x_{i-1}, x_i), \; 1\leq i \leq k+1,$   according to the following formulas
\[
R(1)=2\Delta_1-4\frac{z_1}{M},
\]
\begin{equation}\label{eq:14}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\frac{z_i+z_{i-1}}{M},1<i<k+1,
\end{equation}
\[
R(k+1)=2\Delta_{k+1}-4\frac{z_k}{M}.
\]

Step 4. Find interval $(x_{t-1},x_t)$ with maximum characteristic
\[
R(t)=\max{\left\{R(i): 1 \leq i \leq k+1\right\}}.
\]

Step 5. Carry out a trial in point $x^{k+1}\in(x_{t-1},x_t)$, calculated using the following formulas
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2}, t=1, t=k+1,
\]
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N, 1<t<k+1.
\]

The algorithm stops working if the condition $\Delta_t<\epsilon$ is satisfied; here $\epsilon>0$ the accuracy is preset. For estimation of the global solution values
\[
f_k^\ast=\min_{1\leq i \leq k}f(x^i), \ x_k^\ast=\arg \min_{1\leq i \leq k}f(x^i).
\]
are selected.

The proof of convergence of this algorithm is provided in \cite{RefStrongin2000}. The modifications considering existence of inequality constraints in the problem and information about objective function derivative are given in \cite{RefBarkalov,RefGergel1996,RefGergel1997}.

Let us illustrate the work of this \textit{global search algorithm} (GSA) using a dimension reduction for minimization of a two-dimensional multiextremal function
\begin{eqnarray} \nonumber \label{eq:19}
\varphi(y)= -&\left\{\left(\sum^{7}_{i=1}\sum^{7}_{j=1}A_{ij}g_{ij}(y)+B_{ij}h_{ij}(y)\right)^2+\right. \\
&\left.\left(\sum^{7}_{i=1}\sum^{7}_{j=1}C_{ij}g_{ij}(y)+D_{ij}h_{ij}(y)\right)^2\right\}^{1/2},\\ \nonumber
\end{eqnarray}
where
\begin{eqnarray} \nonumber
& y=(y_1,y_2)\in R^2, 0 \leq y_s \leq 1, s=1,2, \\ \nonumber
& g_{ij}(y)=\sin(i\pi y_1)\sin(j\pi y_2),  \\ \nonumber
& h_{ij}(y)=\cos(i\pi y_1)\cos(j\pi y_2), \nonumber 
\end{eqnarray}
and coefficients $A_{ij}, B_{ij}, C_{ij}, D_{ij}$  are taken uniformly in interval $[-1,1]$. The experiment used method parameters $r=3$, $\epsilon=10^{-3}$, and evolvent construction parameter $m=12$. Fig.~\ref{fig:1} shows contour lines of one of functions generated according to (\ref{eq:19}) and points of 1103 trials carried out using the method before the required accuracy was obtained.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig1.jpg} 
  \caption{Minimization of two-dimensional function}
  \label{fig:1}       % Give a unique label
\end{figure}

@@

First, it is possible to parallelize calculation of the target function describing the optimized object. This method can provide speedup, but is specific for each solved problem.

Secondly, it is possible to parallelize implementation of algorithm computing rules providing selection of the point of the next trial. In this case the way of parallelization will depend on the class of algorithms and, besides, these rules are often rather simple and it is inexpedient to parallelize them (overhead costs of the organization of parallelism will reduce possible speedup to zero).

Thirdly, it is possible to change the scheme of algorithm for the purpose of carrying out several trials in parallel. This approach is the most promising since it is characterized by efficiency (the part of computing process, in which the bulk of calculations is carried out, is parallelized) and generality (it is applicable for a wide class of characteristical algorithms of multiextremal optimization \cite{RefGrishagin1997}).

Let us describe a change of rules of global search algorithm for the purpose of simultaneous (synchronous) carrying out in one iteration of method $p$ of trials. Let us make $k(n)$ the total number of trials, carried out after $n$ of parallel iterations.

Suppose $n\geq 1$  iterations of the method are carried out, in which trials were performed in $k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next $n+1$ iteration are determined according to the rules.

Steps 1--3 of the parallel algorithm fully repeat steps 1--3 of the global search algorithm.

@@

Step 5. Carry out new trials in points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using formulas
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2}, t_j=1, t_j=k+1,
\]
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu}\right]^N, 1<t_j<k+1.
\]

The algorithm stops working if the condition $\Delta_{t_j}<\epsilon$ is satisfied at least for one number $t_j, 1 \leq j \leq p$ ; $\epsilon>0$ is the preset accuracy.  For estimation of the global solution values
\[
f_k^\ast=\min_{1\leq i \leq k}f(x^i), \ x_k^\ast=\arg \min_{1\leq i \leq k}f(x^i).
\]
are selected.

The modifications of the parallel algorithm considering information about the minimized function derivative are given in \cite{RefGergel1999}.

This method of organizing parallel computing has the following justification \cite{RefStrongin2000,RefGrishagin1997}. The characteristics of intervals (\ref{eq:14}) used in the algorithm of global search can be considered as probability measures of location of the global minimum point in these intervals. Inequalities (\ref{eq:21}) arrange intervals according to their characteristics, and trials are carried out in parallel in the first $p$ intervals with the largest probabilities.

Let us use a parallel algorithm of global search for solution of a problem from section \ref{sec:2} with numbers of parallel trials $p=2$ and $p=4$. With $p=2$ the number of iterations was $569$, and the number of trials was $1138$. With $p=4$ the number of iterations was reduced to $270$, and $1080$ trials were carried out. Contour lines of the function are shown in Fig.~\ref{fig:2} with points of the trials carried out by the method when using $p=4$. Comparison of Fig.~\ref{fig:1} and Fig.~\ref{fig:2} shows absence of redundant trial points.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig2.jpg} 
  \caption{Minimization of a two-dimensional function by a parallel algorithm}
  \label{fig:2}
\end{figure}

@@

Let us describe theoretical properties of a parallel algorithm, which characterize its speedup. One of the main indicators of efficiency of parallel algorithms (in any domain, not only in global optimization) is speedup in time
\[
S(p)=T(1)/T(p),
\]
where $T(1)$ is the time spent for solution of a problem by a sequential algorithm, and $T(p)$ is the time of solution of the same problem by a parallel algorithm in a system with $p$ computing elements. The characteristic of efficiency of parallel algorithms (in relation to algorithms of optimization) is also speedup in number of iterations
\begin{equation}\label{eq:26}
s(p)=n(1)p/n(p),
\end{equation}
where $n(1)$ is the number of the trials carried out using the sequential method, and $n(p)$ is the number of the trials carried out using the parallel method with p processors. This characteristic is especially important since for solution of applied problems the time of carrying out of a trial exceeds the time of processing of its results.

It is obvious that number of trials $n(p)$ for sequential and parallel algorithms described in sections \ref{sec:2} and \ref{sec:3} will differ. Actually, the sequential algorithm of global search when selecting point $x^{k+1}$ of the next $k+1$ trial possesses complete information received at the previous $k$ iterations. The parallel algorithm of global search selects not one, but $p$ points $x^{k+j}, 1\leq j \leq p$,  at iteration $k+1$ based on the same information. It means that selection of point $x^{k+j}$ is carried out in absence of information on the trial results in points $x^{k+i}, 1\leq i<j$. Only the first point $x^{k+1}$ will coincide with the point selected by the sequential algorithm. Points of the other trials generally can not coincide with the points generated by the sequential algorithm. Carrying out of such trials can reduce efficiency of use of parallel processors. Therefore, let us consider such trials as ``redundant'', and the value 
\[
\lambda(p) = \left\{
   \begin{array}{lr}
     (n(p)-n(1))/n(p), & n(p) > n(1),\\
     0, & n(p)\leq n(1),
   \end{array}
\right.
\]
as ``method redundancy''.

Let us set the series of trials $\left\{x^k\right\}$ and $\left\{y^m\right\}$ generated correspondingly by sequential and parallel algorithms for solution of the same problem with $\epsilon = 0$ in the condition of stopping. The following theorem from \cite{RefStrongin2000} determines the number of computing elements $p$, which can be involved for non-redundant parallelization.

\textbf{Theorem}. Suppose $x^\ast$ is the point of global minimum, $x'$ is the point of the local minimum of function $f(x)$, and the following conditions are fulfilled:
\begin{enumerate}
	\item
	Inequality 
	\begin{equation}\label{eq:28}
	f(x')-f(x^\ast)\leq \delta, \delta>0,
	\end{equation}
	holds.
	\item
	The initial $q(l)$ trials of the sequential and parallel methods coincide, i.e.
	\[
	\left\{x^1,\dots,x^{q(l)}\right\}=\left\{y^1,\dots,y^{q(l)}\right\},
	\]
	where 
	\[
	\left\{x^1,\dots,x^{q(l)}\right\}\subset\left\{x^k\right\}, \ \left\{y^1,\dots,y^{q(l)}\right\}\subset\left\{y^m\right\}.
	\]
	\item
	There exists a point $y^n\in \left\{y^m\right\}, n<q(l),$ such that $x'\leq y^n \leq x^\ast$ or \mbox{$x^\ast \leq y^n \leq x'$}.
	\item
	For value $M$ from (\ref{eq:11}) the following inequality
	\[
	M>2^{2-1/N}H
	\]
	holds, where $H$ is H{\"o}lder constant of the minimized function.
\end{enumerate}
